Error Reporting Parsers: a Monad Transformer Approach

### 2013-07-08 article.tex
Edits
Signed-off-by: Edward Z. Yang <ezyang@mit.edu>

\EZY{It's probably worth rephrasing what is stated in the abstract here,
for those of us who are in the habit of skipping abstracts; it also helps
build continuity.}
\Matt{Could you give me any pointers on how to do this?  I can't figure out
an effective way of doing it without just repeating the abstract verbatim.}
\EZY{Hmm, I'd probably convert the abstract into the intro, and write a new,
shorter abstract, like ``Monad transformers can be used to ...''}

@@

\subsection{Parser datatype}
\EZY{These sections are very terse.  If the reader is expected to already be familiar with
most of the moving bits of a parsing library, perhaps this section can be boiled down some more.
It is a very curious audience the current presentation targets: a reader who knows how to
implement a parser library ``from scratch'', but doesn't know how to use
monad transformers to build them.  Perhaps the components of this
presentation which are key to understanding how the monad transformer works should be
fleshed out more.}
\Matt{Good point, I did not have a specific audience in mind.  
The article requires familiarity with monad transformers, but I don't think it should 
require familiarity with parser combinators.  I'll work on improving this section with
that in mind.}

@@

But this ignores the possibility of failure.  To allow for failure, the return value needs to
be wrapped in a Maybe:
\begin{verbatim}
newtype Parser t = 
    Parser {getParser :: [t] -> Maybe [t]}
\end{verbatim}

But we also want parsers to return a value, if they succeed.  So let's extend the return type
again to allow for a value, which also requires a second type parameter:
\begin{verbatim}
newtype Parser t a = 
    Parser {getParser :: [t] -> Maybe ([t], a)}
\end{verbatim}

We've used \verb+newtype+ instead of \verb+type+ so that we can create type class instances
for our parsers later on.

@@

\subsection{Type class instances}
For running multiple parsers in sequence, the combinators from the Monad and
Applicative type classes are great -- which means we need instances for
them.

@@

*Main> getParser parseNest "(()abc(def)()"
Just ("(()abc(def)()",[])
\end{verbatim}
In the rest of this article, we'll rework and augment these basic parser 
combinators by generalizing the implementations and extending the result types,
while taking advantage of monad transformers to keep them modular.

@@

There are two major benefits to our parsers from using monad transformers:
\begin{itemize}
 \item semantic building blocks -- the mtl library provides implementations
    for Maybe, Error, and State building blocks; using these, we can describe
    describe the semantics of our parsers
 \item type class instances -- in addition to providing the building blocks,
    the mtl provides instances for common type classes, so we only have to 
    worry about parser-specific operations
\end{itemize}

>>>>

@@

\subsection{Parser datatype}

@@

We also want parsers to return a value, if they succeed.  So let's extend the return type
again to allow for a value, which also requires a second type parameter:
\begin{verbatim}
newtype Parser t a = 
    Parser {getParser :: [t] -> ([t], a)}
\end{verbatim}

However, this ignores the possibility of failure.  To allow for failure, the return value needs to
be wrapped in a Maybe:
\begin{verbatim}
newtype Parser t = 
    Parser {getParser :: [t] -> Maybe ([t], a)}
\end{verbatim}

This gives us a simple, minimal definition of a parser.  We've used
\verb+newtype+ instead of \verb+type+ so that we can create type class
instances for our parsers later on.

@@

\subsection{Type class instances}
For running multiple parsers in sequence, the combinators from the Monad and
Applicative type classes are very useful; our next step is to implement them.

@@

*Main> getParser parseNest "(()abc(def)()"
Just ("(()abc(def)()",[])
\end{verbatim}
That concludes the implementation of a simple parser.
In the rest of this article, we'll rework and augment these basic parser 
combinators by generalizing the implementations and extending the result types,
while taking advantage of monad transformers to keep them modular.

@@

There are two major benefits to our parsers from using monad transformers:
\begin{itemize}
 \item semantic building blocks -- the mtl library provides implementations
    for Maybe, Error, and State building blocks; using these, we can describe
    describe the semantics of our parsers, and
 \item type class instances -- in addition to providing the building blocks,
    the mtl provides instances for common type classes, so we only have to 
    worry about parser-specific operations.
\end{itemize}
